Computing properties of independent sets by generic programming tensor networks \thanks{\funding{...}}

### 2021-12-27 paper/paper.tex
fix xungao comments

\begin{abstract}
We introduce a method of using generic programming tensor network to compute various properties of independent sets,
which includes: maximum size of such sets, the number of sets of a given size, enumeration of sets of a given size.
%Although the computational complexity inevitably scales exponentially with treewidth of the graph,
%our method using tensor network contraction with generic programming is highly versatile and is able to compute properties that are otherwise not known to be possible for up to several hundred vertices.
At the same time, our algorithms are very simple to implement and one can directly utilize recent advances in tensor network contraction techniques like near optimal contraction order finding and slicing.
%One can easily find a good contraction order for the tensor network representation of a graph $G$ with thousands of vertices. 
The algorithmic complexity of this approach is $2^{{\rm tw}(G)}$, where ${\rm tw} (G)$ the tree width of the problem graph.
Our generic programming tensor network method can be used to compute properties of other problems like cut size, coloring, maximal cliques et al.
To demonstrate the versatility of this tool, we apply it to a few examples including the calculations of the hard-square entropy constant, Euler characteristics of the independence complex, partition functions, and finite-temperature phase transitions on square-lattice graphs\blue{remove some examples}.
\end{abstract}

@@

\blue{independence polynomial at -1~\cite{Bousquet2008,Levit2009},
two motivations of computing independence polynomial: 1. Lee-Yang zero~\cite{Lee1952,Yang1952}, 2. Lov{\'a}sz local lemma~\cite{Scott2005}.
}
\section{Introduction}
In graph theory and combinatorial optimization, there are many interesting hard computational problems concerning properties of independent sets.
%An independent set of a graph is a subset of its vertices that no two vertices in this set are adjacent to each other.
For an undirected graph $G = (V,E)$, an independent set $I \subseteq V$ is a set of vertices that for any vertex pair $u,v \in I$, $(u, v) \not\in E$. %there is no edge connecting $u$ and $v$ in $G$.
The primary insterest of many computational complexity theorists is finding the maximum size of such vertex sets $\alpha(G) \equiv \max_{I}|I|$ and one of the sets of this size.
Finding $\alpha(G)$ is a hard computational problem; moreover, it is NP-hard to even approximate $\alpha(G)$ within a factor $|V|^{1-\epsilon}$ for an arbitrarily small positive $\epsilon$ ~\cite{Hastad1996}.
Naive exhaustive search for an MIS requires computing time $O\left(2^{|V|} \right)$.
People developed more efficient exact algorithms for finding MISs, including the branching algorithms~\cite{Tarjan1977,Robson1986} and dynamic programming.
The branching algorithms can reduce the base of the exponential time scaling to, e.g., $1.1996^{|V|} {|V|}^{O(1)}$~\cite{Xiao2017}, 
while dynamic programming approaches~\cite{Courcelle1990, Fomin2013} works better for graphs with a small treewidth $\text{tw}(G)$;
these methods can produce algorithms of complexity $O(2^{\text{tw}(G)} \text{tw}(G) |V|)$.
People are eager to find better algorithms for solving the MIS problem;
not only for it has a wide range of applications in scheduling, logistics, social network analysis, bioinformatics, wireless networks and telecommunication,
map labelling, computer vision, etc.~\cite{Butenko2003, Wu2015}, but also for it is a well know NP-complete problem that can be mapped from
 to many other important combinatorial optimization problems in polynomial time such as circuit satisfiability~\cite{}, the maximum clique problem and the maximum vertex cover~\cite{Moore2011} problems.

%more concretely, an maximal independent set of a graph $G$ corresponds to a maximum clique in the complement graph of $G$,
%and the complement vertex set of an MIS corresponds to a minimum vertex cover of the graph $G$.
%In this paper, the term ``maximal'' has a different meaning from ``maximum''  which we discussed above; a maximal independent set is an independent set that is not a subset of any other independent set, but its size may not be the maximum.
In this paper, we do not limit our discussion to finding one of the maximum indepedent sets and its size.
There are many other interesting problems pertaining to independent sets, such as the number of independent sets at a given size
and enumerating independent sets at a give size or what we call computing \textit{properties}.
Some of these problems are of great interest in physics applications such as the hard-core lattice gas model~\cite{Dyre2016, Fernandes2007}
in statistical mechanics and the Rydberg hamiltonian with neutral atoms~\cite{Pichler2018} \red{cite experiment when it's ready};
they can, for example, be used to understand phase transitions, to analyse the overlap gap properties~\cite{Gamarnik2013, Gamarnik2019} and to identify harder graphs in an ensemble of graphs \red{cite experiment}.
The problem of computing the independence polynomial belongs to the complexity class \#P-hard. One approach to exactly compute the independence polynomial requires a computational time of $O(1.442^{|V|})$~\cite{Ferrin2014}
\blue{I am not sure about this complexity, this is based on the naive analysis of theorem 2.2 in ~\cite{Ferrin2014}}. There are some interesting works on efficiently approximating the independence polynomial~\cite{Harvey2018}, but in this work, we focus on computing the independence polynomial exactly.
We introduce a tensor-network based framework to compute the various properties pertaining to independent sets.
We map the independent set problem into generic tensor network contraction with specially designed tensor element algebra.
Its algorithmic complexity is similar to that of dynamic programming, i.e. scales exponentially as ${\rm tw}(G)$.
%But the tensor network approach is highly versatile so different problems can be generically solved with minimal implementation efforts without loss of efficiency.
We benchmark our algorithms by computing various independent set properties on central processing units (CPUs) and graphics processing units (GPUs) and show its good performance on sparse graphs.
Its high performance benefits from recent advances in random tensor-network contraction for the purposes of quantum circuit simulations~\cite{Gray2021, Pan2021, Kalachev2021}.
People developed algorithms to optimize the contraction order for tensor networks with thousands of vertices,
and the optimized contraction order gives us a near optimal space complexity $2^{\sim{\rm tw}(G)}$ to contract the tensor network.
For cases when the tensors being too large to fit into a GPU, slicing technique can be used to further reduce the space consumption.
Lastly, we provide a few examples and show that this toolbox can be used to calculate the hard-square entropy constant,
Euler characteristics of the independence complex, partition functions, and finite-temperature phase transitions on square-lattice graphs. 
Our method can also be used for finding ``maximal'' independent set properties.
Here, the term ``maximal'' has a different meaning from ``maximum''  which we discussed above;
a maximal independent set is an independent set that is not a subset of any other independent set, but its size may not be the maximum.
We show how to compute properties related to maximal independent sets as well as other combinatorial optimization problems in \App{app:otherproblems}.
The implementation of our algorithm is publicly available online: ~\cite{repolink}.

@@

\begin{definition}[Generic programming]
   Generic programming is an approach to programming that focuses on designing algorithms and data structures so that they work in the most general setting without loss of efficiency. ~\cite{Stepanov2014}
\end{definition}

@@

\section{Independence polynomial}
\subsection{Independence polynomial}\label{sec:indpoly}
The independence polynomial is an important graph polynomial that contains the counting information of independent sets. It is defined as
\begin{equation}\label{eq:idpdef}
I(G, x) = \sum_{k=0}^{\alpha(G)} a_k x^k,
\end{equation}
where $a_k$ is the number of independent sets of size $k$ in $G$. The total number of independent sets is thus equal to $I(G, 1)$.
To compute the independence polynomial of a graph $G = (V, E)$, we reinterpret this problem as a tensor network contraction problem.
We map a vertex $i\in V$ to a label $s_i \in \{0, 1\}$ of dimension $2$ in a tensor network, where we use $0$ ($1$) to denote a vertex absent (exists) in the set.
For each label $s_i$, we defined a prametrized rank-one vertex tensor $W(x_i)$ indexed by it as
\begin{equation}
    W(x_i) = \left(\begin{matrix}
        1 \\
        x_i
    \end{matrix}\right).
\end{equation}
On each edge $(u, v)$, we define a matrix $B$ indexed by $(s_u, s_v)$ as
\begin{equation}
    \qquad \quad 
       B = \left(\begin{matrix}
        1  & 1\\
        1 & 0
    \end{matrix}\right), \label{eq:edgetensor}
\end{equation}
The corresponding tensor network contraction gives
\begin{equation}\label{eq:idp}
    P(G, \{x_1, x_{2}, \ldots,x_{|V|}\}) = \sum\limits_{s_1, s_2, \ldots, s_{|V|} = 0}^{1} \prod\limits_{i=1}^{|V|} W(x_i)_{s_i} \prod\limits_{(i,j) \in E(G)} B_{s_i s_j},
\end{equation}
where the summation runs over all vertex configurations $\{s_1, s_{2}, \ldots,s_{|V|}\}$ and accumulates the product of tensor elements to the output $P$ (see Example \ref{eg:tensorcontraction} for a concrete example).
The edge tensor element $B_{s_{i}=1, s_{j}=1} = 0$ encodes the independent set constraint, meaning vertex $i$ and $j$ cannot be both in the independent set if they are connected by an edge $(i,j)$.
In the special case of $x_i = x$, the contraction result directly corresponds to the independence polynomial.
The connection can be understood as follows: the product over vertex tensor elements produces a factor $x^k$, where $k=\sum_i s_i$ counts the set size,
and the product over edge tensor elements gives a factor $1$ for a configuration being in an independent set and $0$ otherwise.

@@

\section{Maximum independent sets and its counting}
\subsection{Tropical algebra for finding the MIS size and counting MISs}
In the previous section, we focused on computing the independence polynomial for a graph $G$ of given MIS size $\alpha(G)$, but we did not show how to compute this number.
The method we use to compute this quantity is based on the following observations. Let $x=\infty$, the independence polynomial becomes
\begin{equation}
I(G, \infty) = a_{\alpha(G)} \infty^{\alpha(G)},
\end{equation}
where the lower-order terms vanish. We can thus replace the polynomial type $a = (a_0, a_1, \ldots, a_k)$ with new type with two fields: the largest exponent $k$ and its coefficient $a_k$.
From this, we can define a new algebra as
%\begin{equation}
%\begin{split}
%    a(k) \oplus a(j) &= \begin{cases}
%        \left(a_k + a_j, \max(k,j) \right), & k = j \\
%        \left(a_j, \max(k,j) \right), & k < j \\
%        \left(a_k, \max(k,j) \right), & k > j
%    \end{cases}, \\
%    a(k) \odot a(j) &= (a_k a_j, k+j) \\
%    \mymathbb{0} &= (0, -\infty)\\
%    \mymathbb{1} &= (1, 0). \label{eq:countingtropical}
%\end{split}
%\end{equation}
\begin{equation}
    \eqname{P1}
\begin{split}
    a_x\infty^x \oplus a_y\infty^y &= \begin{cases}
        (a_x + a_y)\infty^{\max(x,y)}, & x = y\\
        a_y\infty^{\max(x,y)}, & x < y\\
        a_x\infty^{\max(x,y)}, & x > y
    \end{cases}, \\
    a_x\infty^x \odot a_y\infty^y &= a_x a_y\infty^{x+y}\\
    \mymathbb{0} &= 0\infty^{-\infty}\\
    \mymathbb{1} &= 1\infty^{0}.
\end{split}%& \text{\rlap{(P1)}}
\label{eq:countingtropical}
\end{equation}
When we are only interested in knowing the MIS size, we can drop the counting field.
The algebra of the exponents becomes the max-plus tropical algebra~\cite{Maclagan2015, Moore2011}.:
\begin{equation}\eqname{T}
    \begin{split}
        x \oplus y &= \max(x,y)\\
        x \odot y &= x + y\\
        \mymathbb{0} &= -\infty\\
        \mymathbb{1} &= 0.
    \end{split}\label{eq:tropical}
\end{equation}
This algebra is the same as the one used in Liu et al.~\cite{Liu2021} to calculate and count spin glass ground states.
For independent set calculations here, the vertex tensor and edge tensor becomes:
\begin{equation}
    W^{\rm tropical} = \left(\begin{matrix}
        \mymathbb{1} \\
        \infty
    \end{matrix}\right),   
    \qquad \qquad
        B^{\rm tropical} = \left(\begin{matrix}
        \mymathbb{1}  & \mymathbb{1} \\
        \mymathbb{1} & \mymathbb{0}
    \end{matrix}\right).
\end{equation}

@@

\section{Enumeration of independent sets}
\subsection{Set algebra for configuration enumeration}
The enumeration problems of independent sets are also interesting and has been studied extensively in the literature~\cite{Bron1973, Eppstein2010, Johnson1988}, including,
for example, the enumeration of all independent sets, the enumeration of all maximal independent sets, or the enumeration of all MISs.
To enumerate all independent sets, we designed an algebra defined on sets of bitstrings.
\begin{equation}
\eqname{SN}
\begin{split}
    s \oplus t &= s \cup t\\
    s \odot t &= \{\sigma \lor^\circ \tau \, | \, \sigma \in s, \tau \in t\}\\
    \mymathbb{0} &= \{\}\\
    \mymathbb{1} &= \{0^{\otimes |V|}\}.
\end{split}\label{eq:set}
\end{equation}
where $s$ and $t$ are each a set of $|V|$-bit strings and $\lor^\circ$ is the bitwise OR operation over two bit strings.
\begin{example}\label{eg:setalgebra}
    For elements being bit vectors of length $5$, we have the following set algebra
\begin{equation*}
\begin{split}
    &\{00001\} \oplus \{01110, 01000\} = \{01110, 01000\} \oplus \{00001\} = \{00001,01110, 01000\}\\
    &\{00001\} \oplus \{\} = \{00001\}\\
&\\
    &\{00001\} \odot \{01110, 01000\} = \{01110, 01000\} \odot \{00001\} = \{01111, 01001\}\\
    &\{00001\} \odot \{\} = \{\}\\
    &\{00001\} \odot \{00000\} = \{00001\}
\end{split}
\end{equation*}
\end{example}

To enumerate all independent sets, 
we initialize variable $x_{i}$ in the vertex tensor to $x_i = \{\boldsymbol{e}_{i}\}$, where $\boldsymbol{e}_i$ is the standard basis vector of size $|V|$ and having 1 at index $i$.
The vertex and edge tensors are thus
\begin{equation}
    W^{\rm enum}(\{\boldsymbol{e}_{i}\}) = \left(\begin{matrix}
        \mymathbb{1} \\
        \{\boldsymbol{e}_{i}\}
    \end{matrix}\right),   
    \qquad \qquad
        B^{\rm enum} = \left(\begin{matrix}
        \mymathbb{1}  & \mymathbb{1} \\
        \mymathbb{1} & \mymathbb{0}
    \end{matrix}\right).
\end{equation}

This set algebra can serve as the coefficients in \Eq{eq:countingtropical} to enumerate all MISs, \Eq{eq:polynomial} to enumerate independent sets of different sizes,
or \Eq{eq:max2poly} to enumerate all independent sets of size $\alpha(G)$ and $\alpha(G)-1$.
For example, to enumerate only the MISs, with the tropical algebra, we define $s(k) = (s_{k}, k)$, where the coefficients follows the algebra in \Eq{eq:set} and the orders follows the max-plus tropical algebra.
The combined operations become: 
%\begin{equation}
%\begin{split}
%    s(k) \oplus s(j) &= \begin{cases}
%        \left(s_k \cup s_j, \max(k,j) \right), & k = j \\
%        \left(s_j, \max(k,j) \right), & k < j \\
%        \left(s_k, \max(k,j) \right), & k > j
%    \end{cases}, \\
%    s(k) \odot s(j) &= (\{\sigma \lor^\circ \tau \, | \, \sigma \in s_k, \tau \in s_j\}, k+j) \\
%    \mymathbb{0} &= (\{ \}, -\infty)\\
%    \mymathbb{1} &= (\{0^{\otimes |V|}\}, 0). \label{eq:countingtropicalset}
%\end{split}
%\end{equation}
\begin{equation}
\eqname{P1+SN}
\begin{split}
    s_x\infty^x \oplus s_y\infty^y &= \begin{cases}
        (s_x \cup s_y)\infty^{\max(x,y)}, & x = y\\
        s_y\infty^{\max(x,y)}, & x < y\\
        s_x\infty^{\max(x,y)}, & x > y
    \end{cases},\\
    s_x\infty^x \odot s_y\infty^y &= \{\sigma \lor^\circ \tau | \sigma \in s_x, \tau \in s_y\}\infty^{x+y},\\
    \mymathbb{0} &= \{\}\infty^{-\infty},\\
    \mymathbb{1} &= \{0^{\otimes |V|}\}\infty^{0}. \label{eq:countingtropicalset}
\end{split}
\end{equation}
Clearly, the vertex tensor and edge tensor become
\begin{equation}
    W^{\rm MISenum}(\{\boldsymbol{e}_{i}\}) = \left(\begin{matrix}
        \mymathbb{1} \\
        \{\boldsymbol{e}_{i}\}\infty^1
    \end{matrix}\right),   
    \qquad \qquad
        B^{\rm MISenum} = \left(\begin{matrix}
        \mymathbb{1}  & \mymathbb{1} \\
        \mymathbb{1} & \mymathbb{0}
    \end{matrix}\right).
\end{equation}
The contraction of the corresponding tensor network yields an enumeration of all MIS configurations.

@@

\subsubsection{Euler characteristics of independence complex}

\subsubsection{Partition functions and finite-temperature phase transitions}

\subsubsection{Computing the Hamming distance statistics}

@@

\subsection{Max cut problem}
Max cut problem is also known as the boolean spinglass problem.
For a vertex $v\in V$, we define a boolean degree of freedom $s_v\in\{0, 1\}$.
Then the max cut problem can be encoded to tensor networks by mapping an edge $(i,j)\in E$ to an edge matrix labeled by $(s_i, s_j)$
\begin{equation}
    B(x_{\langle i, j\rangle}) = \left(\begin{matrix}
        1 & x_{\langle i, j\rangle}\\
        x_{\langle i, j\rangle} & 1
    \end{matrix}\right).
\end{equation}
Similar to other problems, we can define a polynomial about edges variables by setting $x_{\langle i, j\rangle} = x$,
where its $k$th coefficient is two times the number of configurations of cut size $k$.

@@

We also computed independence polynomial rigorously for these lattices in our Github repo\blue{add link to the repo}.

>>>>

\begin{abstract}
We introduce a method of using generic programming tensor network to compute various properties of independent sets,
which includes: maximum size of such sets, the number of sets of a given size, enumeration of sets of a given size.
%Although the computational complexity inevitably scales exponentially with treewidth of the graph,
%our method using tensor network contraction with generic programming is highly versatile and is able to compute properties that are otherwise not known to be possible for up to several hundred vertices.
At the same time, our algorithms are very simple to implement and one can directly utilize recent advances in tensor network contraction techniques like near optimal contraction order finding and slicing.
%One can easily find a good contraction order for the tensor network representation of a graph $G$ with thousands of vertices. 
The algorithmic complexity of this approach is $2^{{\rm tw}(G)}$, where ${\rm tw} (G)$ the tree width of the problem graph.
Our generic programming tensor network method can be used to compute properties of other problems like cut size, coloring, maximal cliques et al.
To demonstrate the versatility of this tool, we apply it to a few examples including the calculations of the hard-square entropy constant and overlap gap property of three regular graphs.
\end{abstract}

@@

\section{Introduction}
In graph theory and combinatorial optimization, there are many interesting hard computational problems concerning properties of independent sets.
%An independent set of a graph is a subset of its vertices that no two vertices in this set are adjacent to each other.
For an undirected graph $G = (V,E)$, an independent set $I \subseteq V$ is a set of vertices that for any vertex pair $u,v \in I$, $(u, v) \not\in E$. %there is no edge connecting $u$ and $v$ in $G$.
The primary insterest of many computational complexity theorists is finding the maximum size of such vertex sets $\alpha(G) \equiv \max_{I}|I|$ and one of the sets of this size.
Finding $\alpha(G)$ is a hard computational problem; moreover, it is NP-hard to even approximate $\alpha(G)$ within $|V|^{1-\epsilon}$ ~\cite{Hastad1996},
meaning there is unlikely a polynomial time algorithm to find an independent set with size $\alpha(G)/|V|^{1-\epsilon}$ for an arbitrarily small positive $\epsilon$.
Naive exhaustive search for an MIS requires computing time $O\left(2^{|V|} \right)$.
People developed more efficient exact algorithms for finding MISs, including the branching algorithms~\cite{Tarjan1977,Robson1986} and dynamic programming.
The branching algorithms can reduce the base of the exponential time scaling to, e.g., $1.1996^{|V|} {|V|}^{O(1)}$~\cite{Xiao2017}, 
while dynamic programming approaches~\cite{Courcelle1990, Fomin2013} works better for graphs with a small treewidth $\text{tw}(G)$;
these methods can produce algorithms of complexity $O(2^{\text{tw}(G)} \text{tw}(G) |V|)$.
People are eager to find better algorithms for solving the MIS problem;
not only for it has a wide range of applications in scheduling, logistics, social network analysis, bioinformatics, wireless networks and telecommunication,
map labelling, computer vision, etc.~\cite{Butenko2003, Wu2015}, but also for it is a well know NP-complete problem that can be mapped from
 many other important combinatorial optimization problems in polynomial time such as 3-satisfiability problem, the maximum clique problem and the maximum vertex cover problem~\cite{Moore2011}.

%more concretely, an maximal independent set of a graph $G$ corresponds to a maximum clique in the complement graph of $G$,
%and the complement vertex set of an MIS corresponds to a minimum vertex cover of the graph $G$.
%In this paper, the term ``maximal'' has a different meaning from ``maximum''  which we discussed above; a maximal independent set is an independent set that is not a subset of any other independent set, but its size may not be the maximum.
In this paper, we do not limit our discussion to finding one of the maximum indepedent sets and its size.
There are many other interesting problems pertaining to independent sets, such as the number of independent sets of a given size
and enumerating independent sets at a give size or what we call computing \textit{properties}.
Some of these problems are of great interest in physics applications such as the hard-core lattice gas model~\cite{Dyre2016, Fernandes2007}
in statistical mechanics and the Rydberg hamiltonian with neutral atoms~\cite{Pichler2018} \red{cite experiment when it's ready};
they can, for example, be used to understand phase transitions, to analyse the overlap gap property~\cite{Gamarnik2013, Gamarnik2019} and to identify harder graphs in an ensemble of graphs \red{cite experiment}.
Finding the number of independent sets of a given size is equivalent to computing the coefficients of independence polynomial,
which is a useful graph characteristics that can related to Euler characteristics of the independence complex~\cite{Bousquet2008,Levit2009}, partition functions~\cite{Lee1952,Yang1952} et al.
Computing independence polynomials belongs to the complexity class \#P-hard 
and only exponential algorithms~\cite{Ferrin2014} are known to compute it exactly.
There are also some interesting works on efficiently approximating the independence polynomial~\cite{Harvey2018}, but in this work, we mainly focus on computing the independence polynomial exactly.
We introduce a tensor-network based framework to compute the various properties pertaining to independent sets.
We map the independent set problem into generic tensor network contraction with specially designed tensor element algebra.
Its algorithmic complexity is similar to that of dynamic programming, i.e. scales exponentially as ${\rm tw}(G)$.
%But the tensor network approach is highly versatile so different problems can be generically solved with minimal implementation efforts without loss of efficiency.
We benchmark our algorithms by computing various independent set properties on central processing units (CPUs) and graphics processing units (GPUs) and show its good performance on sparse graphs.
Its high performance benefits from recent advances in random tensor-network contraction for the purposes of quantum circuit simulations~\cite{Gray2021, Pan2021, Kalachev2021}.
People developed algorithms to optimize the contraction order for tensor networks with thousands of vertices,
and the optimized contraction order gives us a near optimal space complexity $2^{\sim{\rm tw}(G)}$ to contract the tensor network.
For cases when the tensors being too large to fit into a GPU, slicing technique can be used to further reduce the space consumption.
Lastly, we provide a few examples and show that this toolbox can be used to calculate the hard-square entropy constant and overlap gap property. 
Our method can also be used for finding ``maximal'' independent set properties.
Here, the term ``maximal'' has a different meaning from ``maximum''  which we discussed above;
a maximal independent set is an independent set that is not a subset of any other independent set, but its size may not be the maximum.
We show how to compute properties related to maximal independent sets as well as other combinatorial optimization problems in \App{app:otherproblems}.
The implementation of our algorithm is publicly available online: ~\cite{repolink}.

@@

\begin{definition}[Generic programming ~\cite{Stepanov2014}]
   Generic programming is an approach to programming that focuses on designing algorithms and data structures so that they work in the most general setting without loss of efficiency.
\end{definition}

@@

\section{Independence polynomial}
\subsection{Independence polynomial}\label{sec:indpoly}
The independence polynomial is an important graph polynomial that contains the counting information of independent sets. It is defined as
\begin{equation}\label{eq:idpdef}
I(G, x) = \sum_{k=0}^{\alpha(G)} a_k x^k,
\end{equation}
where $a_k$ is the number of independent sets of size $k$ in $G$. The total number of independent sets is thus equal to $I(G, 1)$.
To compute the independence polynomial of a graph $G = (V, E)$, we reinterpret this problem as a tensor network contraction problem.
We map a vertex $i\in V$ to a label $s_i \in \{0, 1\}$ of dimension $2$ in a tensor network, where we use $0$ ($1$) to denote a vertex absent (exists) in the set.
For each label $s_i$, we defined a prametrized rank-one vertex tensor $W(x_i)$ indexed by it as
\begin{equation}
    W(x_i) = \left(\begin{matrix}
        1 \\
        x_i
    \end{matrix}\right).
\end{equation}
We use subscripts to index tensor elements, e.g. $W(x_i)_0=1$ is the first element associated with $s_i=0$ and $W(x_i)_1=x_i$ is the second element associated with to $s_i=1$.
Similarly, on each edge $(u, v)$, we define a matrix $B$ indexed by $s_u$ and $s_v$ as
\begin{equation}
    \qquad \quad 
       B = \left(\begin{matrix}
        1  & 1\\
        1 & 0
    \end{matrix}\right). \label{eq:edgetensor}
\end{equation}
The corresponding tensor network contraction gives
\begin{equation}\label{eq:idp}
    P(G, \{x_1, x_{2}, \ldots,x_{|V|}\}) = \sum\limits_{s_1, s_2, \ldots, s_{|V|} = 0}^{1} \prod\limits_{i=1}^{|V|} W(x_i)_{s_i} \prod\limits_{(i,j) \in E(G)} B_{s_i s_j},
\end{equation}
where the summation runs over all vertex configurations $\{s_1, s_{2}, \ldots,s_{|V|}\}$ and accumulates the product of tensor elements to the output $P$ (see Example \ref{eg:tensorcontraction} for a concrete example).
The edge tensor element $B_{s_{i}=1, s_{j}=1} = 0$ encodes the independent set constraint, meaning vertex $i$ and $j$ cannot be both in the independent set if they are connected by an edge $(i,j)$.
In the special case of $x_i = x$, the contraction result directly corresponds to the independence polynomial.
The connection can be understood as follows: the product over vertex tensor elements produces a factor $x^k$, where $k=\sum_i s_i$ counts the set size,
and the product over edge tensor elements gives a factor $1$ for a configuration being in an independent set and $0$ otherwise.

@@

\section{Maximum independent sets and its counting}
\subsection{Tropical algebra for finding the MIS size and counting MISs}
In the previous section, we focused on computing the independence polynomial for a graph $G$ of given MIS size $\alpha(G)$, but we did not show how to compute this number.
The method we use to compute this quantity is based on the following observations. Let $x=\infty$, the independence polynomial becomes
\begin{equation}
I(G, \infty) = a_{\alpha(G)} \infty^{\alpha(G)},
\end{equation}
where the lower-order terms vanish. We can thus replace the polynomial type $a = (a_0, a_1, \ldots, a_k)$ with new type with two fields: the largest exponent $k$ and its coefficient $a_k$.
From this, we can define a new algebra as
%\begin{equation}
%\begin{split}
%    a(k) \oplus a(j) &= \begin{cases}
%        \left(a_k + a_j, \max(k,j) \right), & k = j \\
%        \left(a_j, \max(k,j) \right), & k < j \\
%        \left(a_k, \max(k,j) \right), & k > j
%    \end{cases}, \\
%    a(k) \odot a(j) &= (a_k a_j, k+j) \\
%    \mymathbb{0} &= (0, -\infty)\\
%    \mymathbb{1} &= (1, 0). \label{eq:countingtropical}
%\end{split}
%\end{equation}
\begin{equation}
    \eqname{P1}
\begin{split}
    a_x\infty^x \oplus a_y\infty^y &= \begin{cases}
        (a_x + a_y)\infty^{\max(x,y)}, & x = y\\
        a_y\infty^{\max(x,y)}, & x < y\\
        a_x\infty^{\max(x,y)}, & x > y
    \end{cases}, \\
    a_x\infty^x \odot a_y\infty^y &= a_x a_y\infty^{x+y}\\
    \mymathbb{0} &= 0\infty^{-\infty}\\
    \mymathbb{1} &= 1\infty^{0}.
\end{split}%& \text{\rlap{(P1)}}
\label{eq:countingtropical}
\end{equation}
To implement this algebra programmatically, we create a data type with two fields $(x, a_x)$ to store the MIS size and its counting,
and define the above operations and constants correspondingly.
When we are only interested in knowing the MIS size, we can drop the counting field.
The algebra of the exponents becomes the max-plus tropical algebra~\cite{Maclagan2015, Moore2011}.:
\begin{equation}\eqname{T}
    \begin{split}
        x \oplus y &= \max(x,y)\\
        x \odot y &= x + y\\
        \mymathbb{0} &= -\infty\\
        \mymathbb{1} &= 0.
    \end{split}\label{eq:tropical}
\end{equation}
This algebra is the same as the one used in Liu et al.~\cite{Liu2021} to calculate and count spin glass ground states.
For independent set calculations here, the vertex tensor and edge tensor becomes:
\begin{equation}
    W^{\rm tropical} = \left(\begin{matrix}
        \mymathbb{1} \\
        \infty
    \end{matrix}\right),   
    \qquad \qquad
        B^{\rm tropical} = \left(\begin{matrix}
        \mymathbb{1}  & \mymathbb{1} \\
        \mymathbb{1} & \mymathbb{0}
    \end{matrix}\right).
\end{equation}

@@

\section{Enumeration of independent sets}
\subsection{Set algebra for configuration enumeration}
The enumeration problems of independent sets are also interesting and has been studied extensively in the literature~\cite{Bron1973, Eppstein2010, Johnson1988}, including,
for example, the enumeration of all independent sets, the enumeration of all maximal independent sets, or the enumeration of all MISs.
To enumerate all independent sets, we designed an algebra defined on sets of bitstrings.
\begin{equation}
\eqname{SN}
\begin{split}
    s \oplus t &= s \cup t\\
    s \odot t &= \{\sigma \lor^\circ \tau \, | \, \sigma \in s, \tau \in t\}\\
    \mymathbb{0} &= \{\}\\
    \mymathbb{1} &= \{0^{\otimes |V|}\}.
\end{split}\label{eq:set}
\end{equation}
where $s$ and $t$ are each a set of $|V|$-bit strings and $\lor^\circ$ is the bitwise OR operation over two bit strings.
\begin{example}\label{eg:setalgebra}
    For elements being bit strings of length $5$, we have the following set algebra
\begin{equation*}
\begin{split}
    &\{00001\} \oplus \{01110, 01000\} = \{01110, 01000\} \oplus \{00001\} = \{00001,01110, 01000\}\\
    &\{00001\} \oplus \{\} = \{00001\}\\
&\\
    &\{00001\} \odot \{01110, 01000\} = \{01110, 01000\} \odot \{00001\} = \{01111, 01001\}\\
    &\{00001\} \odot \{\} = \{\}\\
    &\{00001\} \odot \{00000\} = \{00001\}
\end{split}
\end{equation*}
\end{example}

To enumerate all independent sets, 
we initialize variable $x_{i}$ in the vertex tensor to $x_i = \{\boldsymbol{e}_{i}\}$, where $\boldsymbol{e}_i$ is a basis bit string of size $|V|$ that has only one non-zero value at location $i$.
The vertex and edge tensors are thus
\begin{equation}
    W^{\rm enum}(\{\boldsymbol{e}_{i}\}) = \left(\begin{matrix}
        \mymathbb{1} \\
        \{\boldsymbol{e}_{i}\}
    \end{matrix}\right),   
    \qquad \qquad
        B^{\rm enum} = \left(\begin{matrix}
        \mymathbb{1}  & \mymathbb{1} \\
        \mymathbb{1} & \mymathbb{0}
    \end{matrix}\right).
\end{equation}

This set algebra can serve as the coefficients in \Eq{eq:countingtropical} to enumerate all MISs, \Eq{eq:polynomial} to enumerate independent sets of different sizes,
or \Eq{eq:max2poly} to enumerate all independent sets of size $\alpha(G)$ and $\alpha(G)-1$.
As long as the coefficients in a truncated polynomial forms a commutative semiring, the polynomial itself is a commutative semiring.
For example, to enumerate only the MISs, with the tropical algebra, we define $s_{k}\infty^k$,
where the coefficient follows the algebra in \Eq{eq:set} and the orders follows the max-plus tropical algebra.
The combined operations become: 
%\begin{equation}
%\begin{split}
%    s(k) \oplus s(j) &= \begin{cases}
%        \left(s_k \cup s_j, \max(k,j) \right), & k = j \\
%        \left(s_j, \max(k,j) \right), & k < j \\
%        \left(s_k, \max(k,j) \right), & k > j
%    \end{cases}, \\
%    s(k) \odot s(j) &= (\{\sigma \lor^\circ \tau \, | \, \sigma \in s_k, \tau \in s_j\}, k+j) \\
%    \mymathbb{0} &= (\{ \}, -\infty)\\
%    \mymathbb{1} &= (\{0^{\otimes |V|}\}, 0). \label{eq:countingtropicalset}
%\end{split}
%\end{equation}
\begin{equation}
\eqname{P1+SN}
\begin{split}
    s_x\infty^x \oplus s_y\infty^y &= \begin{cases}
        (s_x \cup s_y)\infty^{\max(x,y)}, & x = y\\
        s_y\infty^{\max(x,y)}, & x < y\\
        s_x\infty^{\max(x,y)}, & x > y
    \end{cases},\\
    s_x\infty^x \odot s_y\infty^y &= \{\sigma \lor^\circ \tau | \sigma \in s_x, \tau \in s_y\}\infty^{x+y},\\
    \mymathbb{0} &= \{\}\infty^{-\infty},\\
    \mymathbb{1} &= \{0^{\otimes |V|}\}\infty^{0}. \label{eq:countingtropicalset}
\end{split}
\end{equation}
Clearly, the vertex tensor and edge tensor become
\begin{equation}
    W^{\rm MISenum}(\{\boldsymbol{e}_{i}\}) = \left(\begin{matrix}
        \mymathbb{1} \\
        \{\boldsymbol{e}_{i}\}\infty^1
    \end{matrix}\right),   
    \qquad \qquad
        B^{\rm MISenum} = \left(\begin{matrix}
        \mymathbb{1}  & \mymathbb{1} \\
        \mymathbb{1} & \mymathbb{0}
    \end{matrix}\right).
\end{equation}
The contraction of the corresponding tensor network yields an enumeration of all MIS configurations.

@@

\subsubsection{Computing the overlap gap property}

@@

\subsection{Max cut problem}
Max cut problem is also known as the boolean spinglass problem.
For a vertex $v\in V$, we define a boolean degree of freedom $s_v\in\{0, 1\}$.
Then the max cut problem can be encoded to tensor networks by mapping an edge $(i,j)\in E$ to an edge matrix labeled by $s_is_j$
\begin{equation}
    B(x_{\langle i, j\rangle}) = \left(\begin{matrix}
        1 & x_{\langle i, j\rangle}\\
        x_{\langle i, j\rangle} & 1
    \end{matrix}\right),
\end{equation}
where variable $x_{\langle i, j\rangle}$ represents a cut on edge $(i, j)$ or a domain wall of an Ising spinglass.
Similar to other problems, we can define a polynomial about edges variables by setting $x_{\langle i, j\rangle} = x$,
where its $k$th coefficient is two times the number of configurations of cut size $k$.

@@

We also computed independence polynomial rigorously for these lattices in our \href{https://github.com/GiggleLiu/NoteOnTropicalMIS/tree/master/data}{Github repo}.
